---
title: "Assignment-2"
author: "Vineeth Kadiyam"
date: "2023-10-01"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Assignemnt Problem Statement

  Universal bank is a young bank growing rapidly in terms of overall customer acquisition. The majority of these customers are liability customers (depositors) with varying sizes of relationship with the bank. The customer base of asset customers (borrowers) is quite small, and the bank is interested in expanding this base rapidly in more loan business. In particular, it wants to explore ways of converting its liability customers to personal loan customers.  
  
  A campaign that the bank ran last year for liability customers showed a healthy conversion rate of over 9% success. This has encouraged the retail marketing department to devise smarter campaigns with better target marketing. The goal is to use k-NN to predict whether a new customer will accept a loan offer. This will serve as the basis for the design of a new campaign.
  
  The file UniversalBank.csv contains data on 5000 customers. The data include customer demographic information (age, income, etc.), the customer’s relationship with the bank (mortgage, securities account, etc.), and the customer response to the last personal loan campaign (Personal Loan). Among these 5000 customers, only 480 (= 9.6%) accepted the personal loan that was offered to them in the earlier campaign.
  
  Partition the data into training (60%) and validation (40%) sets.
  
## Load required libraries

```{r}
library(class)
library(caret)
library(e1071)
library(knitr)
```

### Read the UniversalBank data

```{r}
universal.df <- read.csv("UniversalBank.csv")
dim(universal.df)
t(t(names(universal.df))) # The t function creates a transpose of the data frame
```
### Drop ID and ZIP
```{r}
universal.df <- universal.df[,-c(1,5)]
```

### Transform categorical variables into dummy variables

```{r}
# Only Education needs to be converted to factor

universal.df$Education <- as.factor(universal.df$Education)

```

```{r}
# Now, convert Education to Dummy Variables

groups <- dummyVars(~., data = universal.df) # This creates the dummy groups

universal_m.df <- as.data.frame(predict(groups,universal.df))
```

### Split the data to 60% training and 40 % Validation 

```{r}
set.seed(1)  # Important to ensure that we get the same sample if we rerun the code
train.index <- sample(row.names(universal_m.df), 0.6*dim(universal_m.df)[1])
valid.index <- setdiff(row.names(universal_m.df), train.index)  
train.df <- universal_m.df[train.index,]
valid.df <- universal_m.df[valid.index,]
t(t(names(train.df)))

```

### Now normalize the data

```{r}
train.norm.df <- train.df[,-10] # Note that Personal Income is the 10th variable
valid.norm.df <- valid.df[,-10]

norm.values <- preProcess(train.df[, -10], method=c("center", "scale"))
train.norm.df <- predict(norm.values, train.df[, -10])
valid.norm.df <- predict(norm.values, valid.df[, -10])
```

### Question

    1. Age = 40, Experience = 10, Income = 84, Family = 2, CCAvg = 2, Education_1 = 0, Education_2 = 1, Education_3 = 0, Mortgage = 0, Securities Account = 0, CD Account = 0, Online = 1, and Credit Card = 1. Perform a k-NN classification with all predictors except ID and ZIP code using k = 1. Remember to transform categorical predictors with more than two categories into dummy variables first. Specify the success class as 1 (loan acceptance), and use the default cutoff value of 0.5. How would this customer be classified?

# We have converted all categorical variables to dummy variables

## Let's create a new sample
```{r}

new_customer <- data.frame(
  Age = 40,
  Experience = 10,
  Income = 84,
  Family = 2,
  CCAvg = 2,
  Education.1 = 0,
  Education.2 = 1,
  Education.3 = 0,
  Mortgage = 0,
  Securities.Account = 0,
  CD.Account = 0,
  Online = 1,
  CreditCard = 1)
```

### Normalize the new customer

```{r}
new.cust.norm <- new_customer
new.cust.norm <- predict(norm.values, new.cust.norm)

```

### Now let us predict using K-NN(k- Nearest neighbors)

```{r}

knn.pred1 <- class::knn(train = train.norm.df, 
                       test = new.cust.norm, 
                       cl = train.df$Personal.Loan, k = 1)
knn.pred1

```

    2. What is a choice of k that balances between overfitting and ignoring the predictor information?
### Calculate the accuracy for each value of k    
### Set the range of k values to consider

```{r}

accuracy.df <- data.frame(k = seq(1, 15, 1), overallaccuracy = rep(0, 15)) 
for(i in 1:15) 
  {knn.pred <- class::knn(train = train.norm.df, 
                         test = valid.norm.df, 
                         cl = train.df$Personal.Loan, k = i)
  accuracy.df[i, 2] <- confusionMatrix(knn.pred,as.factor(valid.df$Personal.Loan),positive = "1")$overall[1]
}
which(accuracy.df[,2] == max(accuracy.df[,2])) 
plot(accuracy.df$k,accuracy.df$overallaccuracy, main = "Accuracy Vs K", xlab = "k", ylab = "accuracy", col="purple")

```

  3.Show the confusion matrix for the validation data that results from using the best k.

### Confusion Matrix using best K=3

```{r}

knn.predi <- class::knn(train = train.norm.df,
                         test = valid.norm.df, 
                         cl = train.df$Personal.Loan, k = 3)

confusionMatrix(knn.predi,as.factor(valid.df$Personal.Loan))

```

    4.Consider the following customer: Age = 40, Experience = 10, Income = 84, Family = 2, CCAvg = 2, Education_1 = 0, Education_2 = 1, Education_3 = 0, Mortgage = 0, Securities Account = 0, CD Account = 0, Online = 1 and Credit Card = 1. Classify the customer using the best k.

# Load new customer profile

```{r}
new_customer2<-data.frame(
  Age = 40, 
  Experience = 10, 
  Income = 84, 
  family =2, 
  CCAvg = 2, 
  Education_1 = 0,
  Education_2 = 1, 
  Education_3 = 0, 
  Mortgage = 0, 
  Securities.Account = 0, 
  CDAccount = 0, 
  Online = 1, 
  CreditCard = 1)

```


```{r}

knn.predi1 <- class::knn(train = train.norm.df, 
                       test = new.cust.norm, 
                       cl = train.df$Personal.Loan, k = 3)
knn.predi1
```

## Print the predicted class (1 for loan acceptance, 0 for loan rejection)

```{r}

print("This customer is classified as: Loan Rejected")

```

    5. Repartition the data, this time into training, validation, and test sets (50% : 30% : 20%). Apply the k-NN method with the k chosen above. Compare the confusion matrix of the test set with that of the training and validation sets. Comment on the differences and their reason.
    

### Split the data to 50% training and 30% Validation and 20% Testing 

```{r}
set.seed(1)
Train_Index11 <- sample(row.names(universal_m.df), 0.5*dim(universal_m.df)[1])
Val_Index11 <- sample(setdiff(row.names(universal_m.df),Train_Index11),0.3*dim(universal_m.df)[1])
Test_Index11 <-setdiff(row.names(universal_m.df),union(Train_Index11,Val_Index11))
Train_Data1 <- universal_m.df[Train_Index11,]
Validation_Data1 <- universal_m.df[Val_Index11,]
Test_Data1 <- universal_m.df[Test_Index11,]
```

### Now normalize the data

```{r}
train.norm.df1 <- Train_Data1[,-10]
valid.norm.df1 <- Validation_Data1[,-10]
Test.norm.df1  <-Test_Data1[,-10]

norm.values11 <- preProcess(Train_Data1[, -10], method=c("center", "scale"))
train.norm.df1 <- predict(norm.values11, Train_Data1[,-10])
valid.norm.df1 <- predict(norm.values11, Validation_Data1[,-10])
Test.norm.df1 <-predict(norm.values11,Test_Data1[,-10])

```

### Now let us predict using K-NN(k- Nearest neighbors)

```{r}

validation_knn1 = class::knn(train = train.norm.df1, 
                           test = valid.norm.df1,  
                           cl = Train_Data1$Personal.Loan, 
                           k = 3)

test_knn1 = class::knn(train = train.norm.df1, 
                     test = Test.norm.df1,  
                     cl = Train_Data1$Personal.Loan, 
                     k = 3)

Train_knn1 = class::knn(train = train.norm.df1, 
                     test = train.norm.df1,  
                     cl = Train_Data1$Personal.Loan, 
                     k = 3)
```

### Validation confusion Matrix

```{r}
validation_confusion_matrix = confusionMatrix(validation_knn1, 
                                               as.factor(Validation_Data1$Personal.Loan), 
                                               positive = "1")

validation_confusion_matrix

```
### Test confusion Matrix

```{r}

test_confusion_matrix = confusionMatrix(test_knn1, 
                                         as.factor(Test_Data1$Personal.Loan), 
                                         positive = "1")


test_confusion_matrix

```

```{r}

Training_confusion_matrixs = confusionMatrix(Train_knn1, 
                                               as.factor(Train_Data1$Personal.Loan), 
                                               positive = "1")

Training_confusion_matrixs
```


# Difference

##Test vs.Train:

***Accuracy:*** Train has highest accuracy (0.9772) compared to Test (0.9507).

**Reason:** This is as a result of variations in the evaluation datasets. The dataset for Train might be better balanced or predictible.

***Sensitivity (True Positive Rate):*** Train has higher sensitivity (0.7589) compared to Test (0.5875).

**Reason:** This suggests that Train's approach is more accurate at spotting positive cases (such as loan approvals). It might have a decreased rate of false negatives.


***Specificity (True Negative Rate):*** Train has higher specificity (0.9987) compared to Test (0.99403).

**Reason:** This implies that Train's model performs better at accurately recognizing negative cases (such as loan denials). It might have a lower rate of false positives.


***Positive Predictive Value (Precision):*** Train has a higher positive predictive value (0.9827) compared to Test (0.92157).

**Reason:** Fewer false positive predictions are made thanks to Train's model's increased accuracy in predicting positive cases.

## Train vs.Validation:

**Accuracy:** Train still has a more accuracy (0.9772) compared to Validation (0.958).

***Reason:*** Similar to the comparison with Test, Train may have a more balanced or easier-to-predict dataset.

**Sensitivity (True Positive Rate):** Train has more sensitivity (0.7589) compared to Validation (0.625).

***Reason:*** Train's model is better at correctly identifying positive cases. This indicates that Validation's model may have a higher false negative rate.

**Specificity (True Negative Rate):** Train has more specificity (0.9987) compared to Validation (0.9934).

***Reason:*** Train's model is better at correctly identifying negative cases. Validation's model may have a slightly higher false positive rate.

**Positive Predictive Value (Precision):** Train still has a more positive predictive value (0.9827) compared to Validation (0.9091).

***Reason:*** Train's model is more precise in predicting positive cases, resulting in fewer false positive predictions.

## Potential Reasons for Differences:

**Data set Differences:** Performance of the model can be considerably impacted by differences in the make-up and distribution of the data across distinct sets. As an example, one data collection could be more unbalanced than another, making it more difficult to forecast unusual events.  e unbalanced, making it more difficult to predict unusual events.   

**Model Variability:** Performance variances may be caused by different model configurations or uncontrolled initialization of model parameters.

**Hyperparameter Tuning:** Model performance may be impacted by several hyper parameter settings, including the choice of k in k-NN or other model-specific factors.

**Data unyoking:** Results may vary, especially for small data sets, if the data sets are divided into training, confirmation, and test sets for each evaluation.

**Sample Variability:** Variations in the particular samples included in the confirmation and test sets can have an impact on performance criteria in small data sets.

**Randomness:** Some optimization processes for models, such as those used in neural networks, include randomness, which can produce small fluctuations. 
